BarnebyLives an R package to create herbarium specimen labels

### 2023-09-04 manuscript/manuscript_APPS.rmd
adding sys times

--- 
title: BarnebyLives an R package to create herbarium specimen labels
author:  |
    | Reed Clark Benkendorf$^1$^[Correspondence: rbenkendorf@chicagobotanic.org],  Nyree Zerega$^1$$^,$$^2$,  Jeremie B. Fant$^1$$^,$$^2$
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
abstract:  |  
  **Premise:** Depositing quality specimens to herbaria is a time intensive task. Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated to curatorial tasks and specimen deposition. Despite this, the continual generation of herbaria specimens are essential for research in ecology and evolution. In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks and produce labels.    
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. Census Bureau to provide political jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions. It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations. Optionally the package writes driving directions to sites using Google Maps. Finally the package outputs data in a tabular format for review by the user to accept or confirm changes,      
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.     
keywords: |
  Herbarium, collections, natural history museums, geospatial, automation, R
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
---

Herbaria house billions of specimens worldwide. They serve as an essential component of all taxonomic, and phylogenetic studies as well as an enormous number of ecological studies. Recent innovations in computing, digitization and data distribution, DNA sequencing, and the development of myriad analytic approaches have brought about a resurgence in museum based study. However, concerns regarding the rate of acquisition of new specimens exist...

The submittal of specimens to herbaria is a time consuming process, especially for younger collectors with limited experience. 
While many young collectors, who are capable of using dichotomous keys to identify their collections, exist they face difficulties navigating several aspects of data collection. 
This scenario results in the delay of deposition of many specimens, and undoubtedly the deposition of many specimens at all.
Problems which face young collectors generally include the lack of dedicated time awarded to them at a seasons end to process specimens, and the lack of education on several cartographic and natural history attributes.

@@

Here we showcase the reduction in errors attributtal to BarnebyLives

@@

@@

# ACKNOWLEDGEMENTS

>>>>

--- 
title: BarnebyLives an R package to create herbarium specimen labels
author:  |
    | Reed Clark Benkendorf$^1$^[Correspondence: rbenkendorf@chicagobotanic.org],  Nyree Zerega$^1$$^,$$^2$, Pat Herendeen$^1$$^,$$^2$, Jeremie B. Fant$^1$$^,$$^2$
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
abstract:  |  
  **Premise:** Depositing quality specimens to herbaria is a time intensive task. Many institutions have reduced the amount of funding for herbaria, and universities have reduced the amount of education dedicated to curatorial tasks and specimen deposition. Despite this, the continual generation of herbaria specimens are essential for research in ecology and evolution. In order to faciliate the continued growth of herbaria BarnebyLives was developed as tool to supplement collection notes, perform geographic and, taxonomic informatic processes, enact spell checks and produce labels.    
  **Methods and Results:** BarnebyLives uses geospatial data from the U.S. Census Bureau to provide political jurisdiction information, and data from other sources, including the United States Geological Survey, to supplement collection notes by providing information on abiotic site conditions. It uses inhouse spell checks to verify the spelling of a collection at all taxonomic ranks, the IPNI standard author database to check standard author abbreviations, and the Royal Botanic Garden Kews 'Plants of the World Online' to check for nomenclatural innovations. Optionally the package writes driving directions to sites using Google Maps. Finally the package outputs data in a tabular format for review by the user to accept or confirm changes,      
  **Conclusions:** BarnebyLives provides accurate political and physical information, reduces typos, provides users the most current taxonomic opinions, generates driving directions to sites, and produces aesthetically appealing labels and shipping manifests in a matter of minutes.     
keywords: |
  Herbarium, collections, natural history museum, geospatial, automation, R, software
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/american-journal-of-botany.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
- \usepackage[width=\textwidth]{caption}
- \usepackage{wrapfig}
---

```{r}
knitr::opts_chunk$set(echo=F, warning = F, message = F)
```

Nearly 400 million specimens are housed in herbaria around the world (@thiers2021herbaria).
These specimens were collected with the goal of describing the plant kingdoms taxonomic diversity, and documenting the worlds floristic diversity (@greve2016realising). 
The rate of accessioning new collections to herbaria diminished in the 20th centry as research goals in the biological sciences shifted away from describing, documenting, and understanding earths biodiversity (@prather2004decline, @pyke2010biological).
Which, among other factors, lead to a decline in the amount of funding allocated to collections based research, and the number of staff maintaining and accessioning new collections (@funk2014erosion). 
Fortunately, renewed interest in collections have brought herbaria back to the forefront of plant sciences (@ronsted2020integrative).

Recent innovations in computing, specimen digitization, data sharing, DNA sequencing, and statistics have brought about a renaissance in herbarium based studies (@greve2016realising, @james2018herbarium, @brewer2019factors, @ronsted2020integrative). 
Current uses of specimen based data extend far beyond their traditional roles in systematics and floristics, and studies utilizing collections are regularly carried out to better understand .... of plants (). 
However, we anticipate that collections will gain their most widespread utilization as natural history is being revitalized in ecology, via novel approaches, such as remote sensing, meta-barcoding, and ... (@tosa2021rapid).

However, we now stand at a time where we recognize the need for more specimens, but are in a difficult position where the skills of collecting and processing specimens have declined amongst young persons (Daru et al 2017).
The submittal of specimens to herbaria is a, well documented albeit time consuming process, especially for younger collectors with limited experience in the process. 
While many young collectors, who are capable of using dichotomous keys to reliably identify their collections, exist we have observed that they face difficulties navigating several aspects of data collection. 
This scenario results in not only the delay in the deposition of many specimens, but undoubtedly the deposition of many collections at all.
Problems which young collectors face generally include the lack of dedicated time awarded to them at a seasons end to process specimens, and the lack of formal education on cartography, natural history, and taxonomy at most universities.

@@

@@

```{r load libraries}
library(tidyverse) # data operations
library(sf) # spatial data
library(BarnebyLives) # for helping accession collections
```

```{r import example data and gather some summaries}

test_data <- read.csv('data/test_data.csv', na.strings = "") %>% 
  drop_na(longitude) %>% 
  drop_na(latitude) %>% 
  drop_na(date_digital) %>% 
  st_as_sf(coords = c('longitude', 'latitude'), crs = 4326)

n_families <- test_data %>% 
  st_drop_geometry() %>% 
  group_by(family) %>% 
  count() %>% 
  nrow() # 74 families

n_spp <- test_data %>% 
  st_drop_geometry() %>% 
  group_by(binomial) %>% 
  count() %>% 
  nrow() # 616 species

n_infraspecies <- test_data %>% 
  st_drop_geometry() %>% 
  drop_na(infraspecies) %>% 
  group_by(binomial, infraspecies) %>% 
  count() %>% 
  nrow() # 66 distinct infraspecies

n_sp_authors <- test_data %>% 
  st_drop_geometry() %>% 
  drop_na(authority) %>% 
  count() # 557 groups of authors

n_infra_sp_auths <- test_data %>% 
  st_drop_geometry() %>% 
  drop_na(authority1) %>% 
  group_by(binomial, infraspecies) %>% 
  count() %>% 
  nrow() # 22 distinct infra species author groups
```
```{r Run pipeline with all steps and benchmarking, eval = F}

system.time({ # if neccessary convert coordinates in degrees minutes second to decimal degrees
  data <- dms2dd(data, dms = F)
})

time_coords2sf <- system.time({ # create a spatial (simple features) object
  data <- coords2sf(data)
})

time_political_grabber <- system.time({ # grab political information for collection
  political_grabber(data, y = 'Collection_number', path = p)
})

time_physical_grabber <- system.time({ # grab sites physical information
  data <- physical_grabber(data, path = p)
})

time_site_writer <- system.time({ # write site location notes
  data <- site_writer(data, path = p)
})

time_directions_grabber <- system.time({ # write directions to sites
  SoS_gkey = Sys.getenv("Sos_gkey")
  data <- directions_grabber(data, api_key = SoS_gkey)
})

time_spell_check <- system.time({ # ensure appropriate spellings at taxonomic levels
  data <- spell_check(data, path = p)
})

time_author_check <- system.time({ # ensure authorities are spelled-noted correctly
  data <- author_check(data, path = p)
})

time_powo_searcher <- system.time({ # search for synonyms from plants of the world online

names <- sf::st_drop_geometry(data) %>% 
  pull(Full_name)

pow_res <- lapply(names,
      powo_searcher) %>% 
  bind_rows()
data <- bind_cols(data, pow_res)

rm(names, pow_res)
})

time_associate_dropper <- system.time({ # remove the focal taxon from the noted associates
  data <- associate_dropper(data)
})

time_date_parser <- system.time({ # parse dates into museum formats
  data <- date_parser(input, coll_date = 'Date_digital', det_date = 'Determined_date')
})

time_geodata_writer <- system.time({ # write out collection as GoogleEarth object
  geodata_writer(data, path = 'someplace', 
               filename = 'Herbarium_Collections_2023',
               filetype = 'kml')
})

```

Here we showcase the reduction in errors which are associated with the implementation of BarnebyLives.

@@

# AUTHOR CONTRIBUTIONS
The project was conceptualized by R.C.B. The program was written by R.C.B Data collection and analysis were performed by R.C.B. R.C.B. wrote the manuscript with input from all other authors. All authors approved the final version of the manuscript.

# ACKNOWLEDGEMENTS
The Bureau of Land Management are graciously acknowledged as providers of funding to R.C.B for the majority of his specimen collection activities. Two anonymous peer reviewers who increased the quality of this manuscript are thanked. Co-collectors who collected specimens used in this study are thanked: Dani Yashinovitz, Dakota Becerra, Hannah Lovell, Hubert Szczygiel, Caitlin Miller.
